NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-10-12 Manuscript/manuscriptPS.tex
Update text and figures. Figure updates related to https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/21

\subsection{Ca$^{2+}$ binding affinity in bilayers with negatively charged PS lipids}

Ion binding affinity to PS containing membranes can be most conveniently measured 
from PC:PS lipid mixtures where the lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}. In addition, the ion
binding affinity to such mixtures can be detected using the PC lipid headgroup order parameters,
see section~\ref{electrometerFORmixtures}. As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in POPC:POPS (5:1) mixtures 
with respect to the experiments \cite{roux90} (Fig. \ref{changesWITHCaClPS}), indicating overestimated
calcium binding binding affinity. Only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the order parameter changes
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity,
manifested by the peaks in the density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}),
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
in all the other tested models.

@@

The headgroup order parameters of POPS headgroup measured from POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations with a rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}). The $\beta$-carbon order parameter of
POPS increase with the added CaCl$_2$ in the experiment and in all the tested simulation models,
but simulations significantly overestimated the change. The larger $\alpha$-carbon order
parameter of POPS decrease and the smaller one slightly increase with the added
CaCl$_2$ in the experiment. The changes are again significantly overestimated in
the simulations, however, in this case all simulations predict qualitatively different
behaviour. Notably, the changes of POPS headgroup order parameters are overestimated
also in the CHARMM36/NBfix model where the calcium binding affinity was too low. 
We conclude that the effect of bound ions to the headgroup order parameters of POPS
is not qualitatively reproduced by the tested simulations models. This is in contrast
to previous results for PC headgroup \cite{catte16}, where qualitatively correct reponse to bound ions was
observed despite of significant discrepancies in the headgroup structure without additional ions. 
The response of POPS headgroup order parameters to the bound charge is systematic but
less well understood than the responce of PC headgroups used in the electrometer
concept \cite{seelig87,roux90}. The force field development is necessary to
generate MD simulations that could be used to explain the interactions between
PS headgroup and calcium ions.

>>>>

\subsection{Ca$^{2+}$ binding affinity to bilayers with negatively charged PS lipids}

Ion binding affinity to membranes containing negative charged PS lipids can be
measured by detecteding the PC lipid headgroup order parameters
from POPC:POPS (5:1) mixtures (section~\ref{electrometerFORmixtures}),
where the dehydrated lipid-ion complexes and phase separation are 
not observed~\cite{feigenson86,mattai89,roux90,roux91}.
As expected from the previous study of pure PC lipid
bilayers \cite{catte16}, almost all the tested simulation models overestimate the decrease of
POPC headgroup order parameters as a function of Ca$^{2+}$ concentration in POPC:POPS (5:1) mixtures 
with respect to the experiments \cite{roux90} (Fig. \ref{changesWITHCaClPS}), indicating overestimated
calcium binding binding affinity. Only exception is the CHARMM36 model with the NBfix
interaction employed for calcium~\cite{kim16}, which underestimates the order parameter changes
indicating weaker binding affinity than experiments.
%incorporated in the parameters give by the CHARMM-GUI at the time of running the simulations (January 2018),
Notably, CHARMM36 simulations with NBfix corrections \cite{venable13,kim16} give similar binding affinities of
calcium and sodium to POPC bilayer (see section \ref{CHARMMcalciumNBfix}), in contrast to the experimental 
data~\cite{cevc90,akutsu81,altenbach84}. Therefore, we conclude that the calcium binding affinity,
manifested by the peaks in the density distributions along membrane normal (Fig. \ref{CAdensPCPSmixture}),
is underestimated in CHARMM36 simulations with the NBfix for calcium \cite{kim16} but overestimated 
in all the other tested models.

@@

The headgroup order parameters of POPS measured from POPC:POPS (5:1) mixture
exhibit a strong dependence of CaCl$_2$ with small concentrations and rapid saturation
below 100 mM (Fig. \ref{changesWITHCaClPS}).
The order parameter of POPS $\beta$-carbon increase and the larger $\alpha$-carbon 
decrease with the added CaCl$_2$ in experiments. Slight increase is observed in
the smaller $\alpha$-carbon. All the changes are significantly overestimated in the
tested simulation models, including the CHARMM36 with the NBfix for calcium \cite{kim16},
where the binding affinity was underestimated. 
In addition, different simulation models predict qualitatively different behaviour
for the POPS $\alpha$-carbon order parameters with the added calcium.
For example, both order parameters decrease in Berger simulations but increase
in MacRog simulations, while behaviour in Lipid14/17 and CHARMM36 simulations is
more complicated. This is in contrast to the PC headgroup, where
qualitatively correct reponse to the bound ions is observed
in all simulation models despite of the significant discrepancies in the headgroup
structure without additional ions~\cite{catte16}. Therefore, we conclude that
the improvement of force fields is necessary to correctly describe interactions between
PS headgroup and calcium ions using MD simulations.
%The response of POPS headgroup order parameters to the bound charge is systematic but
%less well understood than the responce of PC headgroups used in the electrometer
%concept \cite{seelig87,roux90}.
